Sparse spectral and $p$-finite element methods on disk-slices and rhombi I: zero boundary conditions

### 2019-03-05 papers/halfdisk/halfdisk.tex
Address the points Sheehan made.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk\sotodo{Explain this}. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for some $N \in \No$: \sotodo{Why bmatrix, not pmatrix?}
\begin{align}
\bigPab_n := \begin{bmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{bmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\vdots \\
		\bigPab_N
	\end{bmatrix}
\end{align}
\sotodo{Why not just make $\bigPab$ infinite for now?}and set \(J_x^{(a,b)}, J_y^{(a,b)}\) as the \(\half (N+1)(N+2) \times \half (N+1)(N+2)\) Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \bigPab(x,y) = x \bigPab(x,y), \quad J_y^{(a,b)} \bigPab(x,y) = y \bigPab(x,y).
\end{align}
Then  \(J_x^{(a,b)}, J_y^{(a,b)}\) have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{bmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{bmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{bmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^x_n &:= \begin{bmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots,N \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots,N, \\ 
\nonumber \\
A^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots,N \\
C^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{bmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots,N.
\end{align}

Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs. In other words, sparsity is not dependent on the domain\sotodo{Huh?}.

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{bmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{bmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each \(n = 0,1,\dots,N\):
\begin{align}
A_n &:= \begin{bmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{bmatrix}
		C^x_n \\
		C^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{bmatrix}
		B^x_n \\
		B^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{bmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

@@

????

@@

We can develop a quadrature rule on the half disk \(\Omega\) in order to evaluate (40) and obtain the coefficient vectors \(\mathbf{f}_n\) as follows.

>>>>

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family\bstodo{Is this OK?}. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: \sotodo{Why bmatrix, not pmatrix?}\bstodo{No reason, I am not familiar with pmatrix}
\begin{align}
\bigPab_n := \begin{bmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{bmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{bmatrix}
\end{align}
\sotodo{Why not just make $\bigPab$ infinite for now?}and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\end{align}
Then $J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{bmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{bmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{bmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{bmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{bmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs, meaning that the sparsity is not limited to the specific half-disk case.

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{bmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{bmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:
\begin{align}
A_n &:= \begin{bmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{bmatrix}
		C^x_n \\
		C^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{bmatrix}
		B^x_n \\
		B^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{bmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{n+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots.
\end{cases}
\end{align}

@@

To obtain the recurrence coefficients for the $\{H_n^{(a,b)}\}$ OPs in (4), we use the Lanczos algorithm. \bstodo{Not sure if this is required / what else to say here}

@@

\bstodo{This section contains repetition / expansion on the quadrature rule section above - not sure if we should remove it or move it?} We can develop a quadrature rule on the half disk \(\Omega\) in order to evaluate (40) and obtain the coefficient vectors \(\mathbf{f}_n\) as follows.
